\documentclass[11pt,a4paper]{article}
\usepackage{lac2014}
\sloppy
\newenvironment{contentsmall}{\small}

% This package contains some stylings for the CFP in its submission form.
% For any further use (e.g. in proceedings) it can safely be commented.
%
% NOTE: This document was written using special characters in their
% unescaped form, which is enabled by default when using fontspec.
% If the document is to be compiled differently, someone should make sure
% that special characters are handled correctly (see e.g. the name of author 2).

\usepackage{ulCFPstyles}

\usepackage{mdwlist}
\usepackage{hyperref}

\title{Version Control In Scholarly Editions}

%see lac2012.sty for how to format multiple authors!
\author
{Urs LISKA \and Jan WARCHOŁ
\\ openLilyLib.org / lilypondblog.org
\\ info@openlilylib.org
}



\begin{document}
\maketitle


\begin{abstract}
\begin{contentsmall}
Plaintext-based tools and version control enable scholars to use very efficient
collaborative workflows, solving many inconveniences that currently appear in
the preparation of scholarly editions and allowing a completely new approach to music
publishing.
This paper is based on the authors' experience with an
edition prepared with LilyPond and \LaTeX, as well as the prospects with
another edition currently in preparation.

\end{contentsmall}
\end{abstract}

\keywords{
\begin{contentsmall}
Music engraving, music publishing, scholarly editions, version control, workflows.
\end{contentsmall}
}

\section{Introduction -- our story}
We recently published a revised edition of the songs of Oskar Fried (1871\,--\,1941)%
\footnote{\url{http://en.wikipedia.org/wiki/Oskar\_Fried}}
\footnote{\url{http://www.sound-rel.de}}.
It was realized with a small team of only two musicologists (Urs Liska and Alexander
Gurdon) and one music engraver (Jan Warchoł).
It soon turned out that the project had a larger scale than expected, mainly for two
-- independent -- reasons: the revision dug up a lot more questions than we had thought of,
and the music's complexity was a real challenge for both the notation program and the 
engraver.

Halfway through the project we decided to start using \emph{versioning} to organize
our files, and to use \LaTeX{} to compile text and scores to a contigious volume.
This turned out to be one of the most influential decisions we made in recent years,
because it made us realize to what extent traditional scholarly workflows can be
improved. While using version control is obvious in software development it applies
\emph{equally} to scholarly editions and engraving scores, and in particular it opens
the door to new \emph{collaborative} workflows.

\section{The Problem}
So, what exactly is wrong with current music editing tools and workflows?\\
Well, several things:

\subsection{You have to be careful}
It's easy to accidentally break something. If you're working in a team,
someone can accidentally overwrite good files with bad ones (without you even noticing!).
This problem is magnified by the opaqueness of binary files and \textsc{gui}
tools: your program is doing something with them, but you cannot see what exactly,
which makes them hard to control.

Usually you can use the “undo” function if something goes wrong,
but what if the mistake was already followed by legitimate changes, perhaps
by a collaborator?

We used to constantly check if we haven't broken something,
and it was \emph{extremely} tedious and time consuming.

\subsection{There's a lot on your head}
Keeping track of what changed in the project, who did what,
and which issues on your (usually long) \textsc{todo} list remain
unsolved, is tedious -- especially when your notes are separated from
the actual content, and you have to jump between files to find
what they refer to.

Also, every time you have a longer break from the project, it takes some
time to get back to speed, because you have to recall what you
(and other team members) were doing.

\subsection{Your workflow isn't flexible}
You cannot input music and make layout adjustments in arbitrary order, because
your notation program may not like it -- and if it doesn't like it,
it may spoil your work.

If the deadline is close and you'd like to distribute work between several people,
you're out of luck, because merging their results afterwards is not straightforward
or outright impossible.

If you work in a team, you have to wait until your colleague finishes his part
and “unlocks” the files before you can edit them yourself.

If you're an editor and have submitted your work to the typesetter, who is now cleaning
it up, possibly with a different \textsc{dtp} program, you can no longer do \emph{anything}
with the document.  For every single typo you spot, you have to ask the typesetter to fix it
for you.

Or maybe you are working on a critical edition and you'd like to create two versions
of a score: a verbatim copy of the original manuscript, and a revised edition.
With current tools, you have to finish the unedited version before starting
\emph{any} work on the revision -- so, if any interesting ideas occur to you
while entering the raw material, you can only write them down as \textsc{todo}
items.

\subsection{Some things are simply painful}
If you have multiple versions of a score (or text), all changes have to be applied
multiple times.  Imagine that you have an urtext edition and a revised edition, or
one score in original and one in a transposed key,
and each one is both in A4 and ebook format.  You spot a typo, and\ldots
have to fix it four times!

If you're in the middle of doing some invasive changes and someone asks you for
a copy of your file, you have to give him a work-in-progress version -- or you
have to keep numerous backup copies of the project, labeled `version1`, `version2`
etc. which clutter your computer and create confusion.

\subsection{Underlying causes}
There are two basic causes of these problems:
\begin{itemize*}
\item it's hard to keep track of all changes in a project,
\item units of work cannot be done in parallel, nor be reordered afterwards.
\end{itemize*}

At the core of the second issue is the fact that when you have two different
copies of a document, there is usually no reasonable way of merging them
(i.e. telling the computer to take the changes introduced in one copy
and apply them on top of the changes from the other copy).

\section{The Solution}

\subsection{Version control}
\emph{Version control} is a bread-and-butter concept for any software developer, but the
interesting point is that it can be made equally fruitful for scholarly work and music
engraving. Version control keeps track of \emph{everything} that is happening during
the development of a project. And in particular it allows an arbitrary number of contributors
to work on an arbitrary number of files \emph{at the same time} without problems.

\subsection{Plain text}
Version control performs its work on plain text files because it relies on the 
line-by-line comparison of files. This restricts the choice of tools, but is also the
foundation of the versioning magic -- conflicts may not arise when two people have edited
the same \emph{file} but only when they have edited the same \emph{line!}

\subsection{What do you need?}
In order to fully benefit from versioned
workflows you need to work with tools processing plain text files%
\footnote{Please note that \textsc{xml} file formats \emph{are} plain text files
but less practical for version control. So while it is possible to use versioning on
OpenOffice or MS Office \textsc{xml} files it isn't very useful in our opinion.}
. From our experience
using LilyPond for engraving scores and \LaTeX{} for typesetting texts is a perfect
combination because these two also offer excellent performance in their respective domains.
All needed tools are available as Open Source software, and most of them can
be used on any operating system.

\subsection{Does it really work?}
Editing plain text files may seem daunting and “nerdy”, so it's a valid question
whether it really works and is worth the trouble. We can assure you it \emph{is} worth
any investment of time because the advantages offered by versioned workflows are fundamental.
We hope to make this obvious in this paper.

Before we switched to version control Jan wrote in a note: \emph{work on one song
at a time; \emph{never} beautify more than 2
songs at once}. This wasn't for any specific theoretical reason -- he simply realized
that he couldn't handle more at once without risking serious problems and conflicts.

After the switch we were working on up to 26 songs in parallel and at different topics
(e.\,g. proof-reading and “beautification”) while being completely confident not to
mess up anything.

Versioning enables collaboration at an incredible pace -- you can work on one document
simultaneously, chat in parallel and update your versions every
few minutes. Actually that's how we finished this paper.


\section{How does it work?}
We're not going to describe everything in detail, as this wouldn't fit in the size
constraints of the current submission.  Rather, we will give you a concise summary;
if you want to learn more about the subject we recommend reading an extensive essay
written by Urs%
\footnote{\url{http://lilypondblog.org/2013/07/plain-text-files-in-music/}}.

\subsection{Version control}
\ldots is almost like magic.  Here's a table of contents from the spellbook,
to give you an idea of the basic concepts involved:

\paragraph{Undo/Redo}
On a very basic level you may understand version control as an infinitely flexible
implementation of \emph{undo/redo} mechanisms. However, in contrast to tools you
know from other programs, this isn't restricted to chronologically undoing steps one by
one. Versioning allows you to inspect, revert, modify or reorder \emph{any} modification
individually -- or in groups.

\paragraph{Project history}
The sum or chain of all changesets in a repository forms its \emph{history}. A user can 
at any time inspect any changeset, revert or modify it, or the whole project can be
temporarily set to any given state. This makes it possible to fully retrace the steps
of a project.

\paragraph{Branching}
In a versioned project it is possible to work in \emph{branches}, which can be
understood as a kind of separate working sessions. Work done on one branch doesn't affect
work on other branches. So multiple topics can be worked on at the same time, while
the main line of the project can always be kept in a consistent state.

\paragraph{Line-by-line comparison}
The foundation of version control is the ability to compare a project directory's (also
known as \emph{repository}) content \emph{line by line}. Therefore files don't have to
be locked for editing because modifications are detected on line level.

\subsection{Plain text files}
Plain text file formats are fundamentally different from binary formats,
and have some superior characteristics.

\paragraph{Mergeability and diffability}
are the key properties of plain text files.  Computers know how to handle text,
can precisely determine the differences between multiple copies of a text file,
and know exactly what to do to reapply a change made in one version of a file
on another, slightly different, version of that file.
They can also display a “diff” that shows these changes in a concise, human-readable way.

\paragraph{Transparency:}
Plain text files reveal every aspect of their content, everything is explicitly written
and readable. Anybody can inspect it and see what has been entered and if and how it is
manually tweaked. 

Binary files are opaque and conceal how they manage your input. They can only be
interpreted by the programs that created them, or by programs that have specific input filters.

\paragraph{Separation of concerns:}
The explicit nature of plain text files provides a clean separation of content, meaning
and appearance. Through the use of explicit commands the content
of a document is clearly marked up semantically while this \emph{meaning} is completely
independent from its visual \emph{appearance}.

\paragraph{Readability and stability:}
Plain text files are human readable, binary files are not. This simple statement has
considerable implications. Plain text files can be recovered to a much greater extent
than binary files after being damaged. And they are (at least conceptually) much better
equipped for evolution of applications and operating systems. Plain text file formats are
usually well documented, and if not they can be reverse engineered with significantly
less expense than binary ones.

\subsection{WYSIWYG vs. compilation}
There is another distinction to make, even if it isn't equally crucial to versioning
as the plain text \emph{storage}: the basic user interface. Many people are used to
\textsc{wysiwyg} interfaces where you edit documents graphically. By contrast typical
plain text based tools are \emph{compiled} systems: they take a plain text \emph{input
file} and compile that to the desired \emph{output file}, typically in
\textsc{pdf} format.

LilyPond and \LaTeX{} both are programs that compile plain text input files.

\paragraph{User experience:}
Editing plain text files is something to get used to: writing commands is less intuitive
than moving items with a mouse to most people. And edits don't have an immediate visual
result before the document is compiled the next time. But these drawbacks are counterbalanced
by the advantages of this approach.

\paragraph{Editor independence:}
Plain text tools separate the tasks of editing, processing and displaying documents,
making it possible to choose different tools for different tasks. Plain text files
can principally be edited with \emph{any} text editor, which opens a wide array of
workflow options, for example editing on smartphones or through web interfaces.

\paragraph{Out-of-the-box results}
The fact that modifications of the content don't have to be accomodated in realtime
allows compiled systems to consider more factors when deciding on a document's layout.
Modifications that cause the page breaking to change can bring a
\textsc{wysiwyg} program down to its knees with large documents. In compiled systems the
next run will simply incorporate the update.

An important side-effect of this is the average quality of the out-of-the-box result
compiled systems provide. Complex scores created by LilyPond are usually readable by
default, while most \textsc{wysiwyg} tools produce output that is often unusable without
numerous manual corrections.

Another side-effect is that compiled systems perform much better when asked for providing
documents with alternative layouts. It is a matter of simply including another stylesheet
to create scores or text documents with other paper format or for different media altogether.

\section{Project structure}

\subsection{Libraries}
Plain text file formats provide ways to reference variables or custom
commands and to store these in files that can be included from the
main files. This way one can create libraries of project-wide code
that make the the project consistent. Style sheets and commands are
maintained on a global level, and any modifications automatically
affect all documents or scores.
We split these includes into project specific tools and such of potentially
common usability. The latter were and are published as open source in a new 
LilyPond snippet repository%
\footnote{\url{https://github.com/openlilylib/snippets}}
and as one \LaTeX{} package%
\footnote{\url{http://ctan.org/pkg/lilyglyphs}}.

\subsection{Programming}
Plain text tools like LilyPond and \LaTeX{} offer a level of programmability that
goes beyond usual scripting interfaces (like e.\,g. VBA). Commands may simply apply
some formatting or perform complex operations on multiple arguments.
Programming is a useful method to enforce, say, typographic consistence, but it can
also be used to “process data”.

One interesting use of these possibilities is to create different output for work and for
publication. Elements can be highlighted through coloring (e.\,g. editorial additions)
or laid out in a completely different manner (e.\,g. the entries for the critical report) to 
be more practical to use while still in the editing process. In text documents it is
possible to lay out complex data like the entries for the critical report in a
user friendly way during development and switch layout to a packed version optimized
for space for publication.

An additional aspect (which we didn't use for the current edition) is the fact that
plain text files can be processed by external programs, so it is straightforward to
write small scripts or extensive tools to perform arbitrary operations on your data.
This may be similar to using \emph{plugins} in the graphical notation programs but
can go much deeper into the core of the document files.

\subsection{Mix music and text}
LilyPond and \LaTeX{} are a perfect match for creating scholarly editions. \LaTeX{}
provides several convenient ways of professionally incorporating scores or music examples,
by now there is \texttt{lilyglyphs}%
\footnote{\url{http://ctan.org/pkg/lilyglyphs}}
available to include \emph{any} notational symbols in continuous text -- and both
can be maintained under version control.

Using \texttt{make} can be a convenient tool to automatically produce a newly compiled
volume when individual scores have changed -- or to enforce unit tests before merging
new work.


\section{Collaboration}\label{sec:collaboration}
The rest of this paper is a rather personal report of our experiences with the Fried
edition, explaining in detail how all this worked for us.  It also contains
some comments on a project we're currently starting, and which is intended to become
a reference project for \emph{crowd engraving}, a more or less unknown concept that is
originarily made possible by version control. We also wrote a number of blog posts on
\emph{Scores of Beauty}%
\footnote{\url{http://lilypondblog.org/category/fried-songs}}
which you may enjoy as additional reading.

\bigskip
\hrule

\bigskip
As explained earlier managing collaboration by email or via shared files isn't really
enjoyable because it slows the team down, seriously blocks workflow options and creates considerable risk
of messing things up. So version control was the way to go for us from one point on.
The distributed nature of our versioning tool of choice, Git%
\footnote{\url{http://git-scm.com}. From now on any term that might be used with use
Git terminology (which partially differs from that of other version control systems.)},
made things really comfortable to work with, and the following sections will give you
an overview of our working experience.

\subsection{Distributed work}
In a \textsl{distributed VCS} (version control system), every collaborator
has a local copy of the complete repository and can work freely on that.
Contacting the repository on the server is only necessary to synchronize by downloading
others' work, merging it with one's own and uploading the result%
\footnote{This is significantly different from systems like \emph{CVS} or \emph{Subversion},
where you need to contact the server in order to add an explicit changeset.}
.

As this merging is performed based on line-by-line comparison chances of running into
conflicting situations are rather small.%
\footnote{During more than a year of collaboration on a 100-page long edition
consisting of over 300 files (totalling about 30 thousand lines of code), we only had one conflict.}
% Urs, did you have any conflicts?
And \emph{if} they should occur there are
straightforward ways to recover from the situation. Therefore it is possible to edit files
\emph{simultaneously}, without any need for “locking” files or similar workflow
restrictions. This and the concept of the history allows multiple editors to work freely
without the risk of stepping on each other's toes.

\subsection{Editor and engraver}
% It is actually a hot topic in scholarly discussion
% to what extent the editor should also be an engraver.
% Scholarly editions usually let the editor only provide
% a model, while an engraver from the publishe prepares the
% score. By contrast commercial publishers usually require
% the editor to prepare a near-printable score document.
Traditionally the tasks of the editor and the engraver were completely separated: the
editor somehow prepared an \emph{engraver's copy}, the engraver engraved a metal plate,
the editor did proof-reading, and the engraver finished the plates. In some contexts
this is still common practice, particularly with scholarly edition institutes, whose
editors often don't do any engraving at all. On the other hand many commercial
publishers expect the editor to deliver (nearly) print-ready files that can be
finished in-house with minimal effort (and investment). However, all of these workflows
share the fact that files (or plates) have to be passed around, with strict
“locking strategies” in order to avoid data loss or conflicting situations. If an editor
notices a wrong note this can't simply be fixed in the score file but instead it requires
an email to the engraver asking him to apply the fix etc.

In our plain text based workflow with versioning this is \emph{completely} different.
Everybody works on the same shared data repository and has full access to any file at any
time%
\footnote{We don't want to say that role-based privileges aren't useful in many cases
but want to stress the principal possibility here.}.
Each collaborator can apply any modification because 
\mbox{\emph{a)} this} won't compromise or lose data and
\mbox{\emph{b)} any} modification is meticulously documented in the changesets.%
\footnote{What's more, this information will remain in the history
of the project \emph{forever}. Also, everyone is able to see what changes
happened before and after each modification.}
Particularly the latter aspect is important for the new workflow experience. 
Every edit can be immediately be seen and reviewed by the others%
\footnote{It should be stressed that with plain text files examining the changes
is very straightforward. Even if your project consists of hundreds of thousands of files,
when you ask Git what changes were introduced in a given edit, you will be told the
exact name of the file, and the number of the line in which the change was made.
In fact, Git will also display how this line looked before and after the modification.},
making it comfortable
to apply any necessary modification or fix. For example, if Urs fixed a wrong pitch Jan
realized that he should have a look at that spot and check for potential regressions
in typography.

One consequence of these possibilities is that the tasks and responsibilities of
(scholarly) editor and engraver can and will blur to some extent. If everybody has
permanent access to the complete data, tasks don't have to be separated as
strictly as formerly; this makes a lot of sense because there usually are considerable
intersections of skillsets. Urs as the scholarly editor \emph{does} have strong sense
for typography, and Jan as the engraver \emph{does} have a sharp eye on details, so
he'll notice issues with the musical text too. In a versioned workflow we both can do
as much of the other's task as we like and time allows, improving efficiency but also
the quality of the result. We think this can have significant impact in today's
discussion of the relation between editor and engraver or editor, edition institute and
(commercial) publisher.

When preparing our edition of Oskar Fried's songs this really boosted our work. Being
able to inspect each other's work, to freely apply edits and not the least to encapsulate
work in branches made us work much more straightforwardly. While we had to be
extremely careful when we were still maintaining the Dropbox folder, we could now just
start working on any given issue. Sharing ideas and reviewing each other's work proved
very inspiring and presumably raised the overall quality of our work. We wrote this
several times, and it may sound like a buzz-phrase, but it's \emph{literally} true:
We can no longer imagine how we could have worked without these tools in our “earlier lives”.

Particularly the last phase of the project was very revealing. We had to do our final
proof-reading round after the scores were finished from the typographical perspective.
(Well, this was partly due to suboptimal planning on our part, but one will have this
situation nearly always to some extent.) Modifying a score that has already been
“beautified” is a very delicate task because any change of the musical text can
mess up the layout. Fortunately LilyPond performed very well in this regard%
\footnote{\url{http://lilypondblog.org/2013/11/modifying-already-beautified-scores/}},
but version control made the process less daunting by orders of magnitude.
Urs applied fixes  himself when he and his fellow musicologist found errors (instead
of having to ask Jan to do so) and wrapped each single edit in an individual \emph{commit}%
\footnote{“Commit” is Git's term for a changeset.},
using the commit message as a convenient place to document his opinion on the necessity of
typographic follow-up fixes. Jan could then inspect the chain of messages so he
immediately knew where to look particularly carefully. This ensured that no unintended
side-effect of edits would go unnoticed, making us quite relaxed as soon as we realized
that it really worked out. This situation even allowed Urs to apply or at least prepare
typographic fixes himself, knowing he didn't irreversibly mess anything up and that Jan
could easily identify and revert any change.
Maybe even more important was the fact that we could wrap all this work in separate branches.
The edition comprises 26 songs, and as all of them were handled in individual branches
we could freely work on whichever song was currently convenient, switching context at
will. Urs created a branch for any given song and added only the relevant commits to it.
When ready with that he handed the torch to Jan who would then check everything and
finally merge this branch to the main line again. This way we \emph{always} had the project
as a whole in a consistent state and were able to juggle our tasks around without and fear
and hesitation.

\subsection{Maintainability after publication}
One more aspect to using plain text tools is also based on the meticulous documentation
of the project history. If the source code is written decently from the start it is
\emph{very} straightforward to maintain a project even after the initial publication.
As each fix or improvement will be stored in a changeset you can apply any changes
without messing up anything. There will always be a detailed record about which
changes have been added since publication, or between any given two prints. This is
particularly interesting with today's printing-on-demand strategies where new “runs”
may be “released” after only a few copies or where you might always download the ”latest
version”.


\subsection{Miscellaneous tools}
Apart from version control and plain text based programs there are several tools that
can be used to enhance the “collaborative experience”.

\paragraph{Version control service providers}
While you can run a version control system on your own server there are several online service providers offering useful tools. In addition to plain data exchange they
provide convenient web interfaces that can simplify your life even more.

Of course you will often want to edit your files in dedicated editing environments, but
it \emph{is} a very nice feature to be able to apply small edits to files through the web 
interface of a service provider. For example you can do a quick fix sitting in an internet
cafe etc.

One particularly useful concept in that context are \emph{forks} and \emph{pull requests}.
They can be used perfectly to organize peer review workflows by allowing work to be
integrated to the main line only after review through people with appropriate privileges.
Read more about this in section \ref{sec:proof-reading-and-peer-review} on page \pageref{sec:proof-reading-and-peer-review}.

There is one aspect to consider, though: you may not want to give your data to an external
company. If you want to benefit from web interface features you can install open source
tools on your personal or institution's server%
\footnote{See \url{https://git.wiki.kernel.org/index.php/Interfaces,\_frontends,\_and\_tools\#Web\_Interfaces} for a list of web interfaces to Git},
maybe missing some of the convenience of the commercial providers but instead having complete
control.

\paragraph{Issue trackers}
In software development projects it is a matter of course to use \emph{issue trackers.}
These tools \emph{keep track} of tasks, issues or bugs, helping you not to lose important
\textsc{todo}s along the way. Of course this is also a good idea for scholarly editions. Issue
trackers can be used to assign issues to people, or categorize and prioritize tasks.
They can be very valuable for relentlessly insisting on perfection.

\paragraph{List service providers}
There are services offering project management through a list metaphor%
\footnote{There are many different providers and concepts, so we only mention the one
\emph{we} used: \url{https://trello.com}}.
They use “cards” as their interface that are arranged on boards one way or another.
Similar to issue trackers you can assign items to people, comment on and discuss them and
finally mark them as “done”.

In contrast to issue trackers these services are usually more “agile” and easy to use, as
they usually display many cards at once and offer realtime collaboration with other
team members, powered by modern web technologies that allow to exchange data without
reloading pages. But the drawback of this agility is that it is also volatile.
If the project requires reliable archiving of past events and issue, list apps usually
aren't first or at least single choice.

\medskip
Apart from these tools there are the options to use good old email as a communication channel, mailing lists and forums. Each solution has their merits and problems, very often both are related to the same issues.


\section{Crowd editing}
In the previous section we've talked about the positive impact that version control
has on the conception of collaborative work. While we were actually two people
working together this concept can easily be extended to a larger number of people
involved. Apart from improving traditional editing workflows this can be used to create a
completely new approach to editing scores which we'd give the buzz-phrase \emph{crowd
editing}. If any number of people can participate in a project under version control
this can easily be exploited to distribute the work among a “crowd” of contributors.
Different from the concept of having distinct roles that tend to blur through the
collaborative approach there can be numerous people doing essentially the same kind
of work in parallel, effectively “clustering” manpower. While we're convinced that
in most cases it won't make sense to rely on the swarm intelligence \emph{completely} (so
you should still have a firm project leadership) and that this won't reduce the
absolute number of hours spent on a project, it can significantly reduce the overall
delivery time by distributing these hours to be done in parallel. We think this can
be a serious boost for many kinds of editing projects where meeting tight deadlines
is one of the main challenges.

\subsection{Music entry}
Music entry is a field that can straightforwardly be distributed among a large number
of contributors. Essentially all you need to make this a reliable process is a usable
communictaion infrastructre to avoid duplicate work.
Appropriate tools for this are assignments that can be done in issue trackers or on
virtual cards like on a Trello board -- email is \emph{very} impractical for this kind
of organization. This way the whole material can be effectively divided through the number
of available contributors.

\subsection{Proof-reading and peer review}\label{sec:proof-reading-and-peer-review}
Version control encourages peer review workflows on all stages of a project's lifecycle.
It is not necessary to wait for peer review until work is more or less finished, but it
can be an integral part of the whole development process instead. A natural place to hook
in is when merging working branches into the main line. For example one can require the
merge to be done by a different collaborator than the one who did the work. For music entry
it would for example be appropriate to enter a certain amount of music and then ask someone
else to review and merge it.
This ensures that only reviewed material goes into the project right from the
start. While this won't make a final proof-reading obsolete it will definitely improve
the overall quality of the result.

Online services like Github or Bitbucket provide \emph{pull requests} as a very convenient
tool for this concept. Whenever a branch has changed one can open a pull request as a
publicly visible announcement of changes one wants to be incorporated into the main line.
The pull request can then be inspected, discussed, modified and finally merged or rejected.

\subsection{“Das Trunkne Lied”}
This year we will prepare performance material for “Das Trunkne Lied” by Oskar Fried%
\footnote{\url{http://lilypondblog.org/2013/06/das-trunkne-lied/}}.
This is a really large score for soloists, choir and orchestra, and it will be a
huge reference project for our new workflow. Of course we intend to document it
and exploit it from a theoretical perspective too.

One noteworthy concept is that we split the whole score in tiny segments, each comprising
music from one part, covering the range of one rehearsal mark. Anybody entering or
reviewing music will be facing a rather small file with only this amount of music.
Clever programming (on the “admin's” part) generates a little score from this segment, so
the editor can always work with a very small amount of music. This makes it much more
manageable than working in the context of a huge score -- and it saves huge amounts of
compilation time.
On the other hand one can always compile a complete part or the full score, and LilyPond
will grab all segments that have already been entered and injects them into an otherwise
complete but empty score.

This working evnironment was realized with a combination of programming within LilyPond files (using the built-in extension languag Scheme) as well as programming Python scripts
that for example generate empty stub files for a whole new instrumental part.

\section{Conclusions}\label{sec:conclusions}

We both think that learning version control was the most rewarding thing we did
in the last five years, and we \emph{wholeheartedly} recommend every knowledge
worker to at least give it a try.

If you're not tech-savvy, you may be intimidated at first -- version control
is, after all, an advanced tool developed originally by software enigineers
for themselves.  But its user-friendliness has increased dramatically over the
last years%
\footnote{And we expect this trend to continue.},
and there are now many tools to make working with it easier, as well as an abundance
of excellent documentation.
In the context of another project we're currently preparing a set of documents with a
hands-on introduction to the concepts from the perspective of musicians%
\footnote{\url{https://github.com/engraving-challenges/main}}.

Similarly, making the switch from \textsc{gui} software to plain text workflows
requires effort and learning a new paradigm, but the investment is well worth it.
The ability to effectively use version control is, in our opinion, enough in itself
to make the switch to plain text tools worthwhile. 
And there \emph{are} some promising projects around intending to streamline the user
experience of editing musical scores with plain text document files. For example the
LilyPond \textsc{ide} Frescobaldi%
\footnote{\url{http://www.frescobaldi.org}}
will soon allow you to edit the source code by graphically manipulating the score.
Being able to export LilyPond scores to MusicXML will add new options of interchanging
work with other users and tools.
And it is intended to integrate Git workflows directly into the score editing program!

Being able to easily experiment, collaborate smoothly and work in a more natural way
significantly changes your attitude towards your work and really opens your mind%
\footnote{\url{http://lilypondblog.org/2013/08/version-control-text-quality-and-creativity}}.

While we're really happy with our overall experience, we \emph{do} know that there are
considerable areas of improvement. But we could use our project to collect and evaluate 
many of them and make up our minds about future projects that can make our tools of choice 
even better.
Some of these could well be realized as crowd funded efforts, and we encourage anybody
interested in new ways of (scholarly) music editing powered by Free Software joining our
crusade.

\section{Acknowledgements}

There are several addressees who should receive our thanks. First of all we have to thank our publisher, \emph{sound-rel}%
\footnote{\url{http://www.sound-rel.de}} for making this whole undertaking possible and
enabling us to work on this fascinating task. We know that they have taken a financial
risk whose outcome is still open.

We would definitely not have managed to finish this production in a satisfactory manner
without the generous support from people on the lilypond-user mailing list%
\footnote{\url{https://lists.gnu.org/mailman/listinfo/lilypond-user}}
and the Stackexchange site%
\footnote{\url{http://tex.stackexchange.com}}. In particular we are indebted to
Thomas Morley, David Nalesnik and David Kastrup for providing invaluable insights and
incredible functions for our project and LilyPond.

And finally we are infinitely grateful for all the people who have voluntarily developed
such great tools as LilyPond, \LaTeX{} and Git.

\end{document}
